Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-24 paper/paper.tex
Remove some metavariables

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
% \mathit{Renaming}\\
% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang \scw{Where does the name come from?}}
  \label{fig:syntax}
\end{figure}

% \begin{figure}[h]
%     \[
%       \begin{array}{lll}
%         \mathit{Curried\ Addition} \\
%         add(n) & := & m \mapsto n + m \\ \\
%         \mathit{Extension ([[xi]])} \\
%         [[(xi .: j) 0]]  & := & 0 \\
%         [[(xi .: j) Suc i]]  & := & [[xi j]] \\ \\

%         \mathit{Extension ([[rho]])} \\
%         [[(rho .: a) 0]]  & := & [[a]] \\
%         [[(rho .: a) Suc i]]  & := & [[rho i]] \\ \\

%         \mathit{Up ([[xi]])} \\
%         [[up xi]] & := & (add(1) \circ [[xi]]) , 0 \\ \\

%         \mathit{Renaming} \\
%         [[var i {xi}]] & := & [[xi i]] \\
%         [[(Set i) {xi}]] & := & [[Set i]] \\
%         [[Void {xi}]] & := & [[Void]]\\
%         [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
%         [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
%         [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
%         [[Bool {xi}]] & := & [[Bool]] \\
%         [[true {xi}]] & := & [[true]] \\
%         [[false {xi}]] & := & [[false]] \\
%         [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
%         [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
%         [[refl {xi}]] & := & [[refl]] \\
%         [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\

%         \mathit{Lookup} \\
%         [[(G ++ A) 0]] & := &  [[A]] \\
%         [[(G ++ A) Suc i]] & := & [[G i]] \\ \\

%         \mathit{Drop} \\
%         [[drop 0 G]] & := & [[G]] \\
%         [[drop Suc i (#G ++ A#)]] & := & [[drop i G]] \\ \\
%       \end{array}
%     \]
%   \caption{Auxiliary Functions over Syntax}
%   \label{fig:auxdef}
% \end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.
\scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper since it is more amenable to
automation (Section~\ref{sec:automation}) and leaves no ambiguity
about variable freeness conditions.

%  \scw{Give an overview of the figure}
% \scw{Define a convention that separates your use of natural numbers. Sometimes they
% are universe levels and sometimes they are de Bruijn indices. Better to have separate
% metavariables for each. (And it is possible to generalize universe
% levels to structures other than natural numbers. Ask Jonathan.} \yl{resolved?}
As a dependent type theory, terms and types are collapsed into the same
syntactic category. The type $[[Set i]]$ represent universe
types and $[[var n]]$ represent de Bruijn variables. While
$[[i]],[[j]],$ and $[[n]]$ are all metavariables representing natural
numbers, we always use $[[n]]$ to represent term variables that
participate in substitution and $[[i]],[[j]]$ to represent universe levels.
Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable.
Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.
\scw{Example of what you can use identity types for? Or an example of
a program that uses J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        % \mathit{UpRen} \\
        % [[( up xi ) 0]] & := & [[0]] \\
        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        % \mathit{Renaming ([[a < xi >]])} \\
        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        % \ldots \\ \\
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

>>>>

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
% \mathit{Renaming}\\
% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang \scw{Where does the name come from?}\yl{H =
    hierarchy though it needs a better name}}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.
% \scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
% \yl{need at least 4 vars for J. removed the unnecessary t and C}
We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper since it is more amenable to
automation (Section~\ref{sec:automation}) and leaves no ambiguity
about variable freeness conditions.

%  \scw{Give an overview of the figure}
% \scw{Define a convention that separates your use of natural numbers. Sometimes they
% are universe levels and sometimes they are de Bruijn indices. Better to have separate
% metavariables for each. (And it is possible to generalize universe
% levels to structures other than natural numbers. Ask Jonathan.} \yl{resolved?}
As a dependent type theory, terms and types are collapsed into the same
syntactic category. The type $[[Set i]]$ represent universe
types and $[[var n]]$ represent de Bruijn variables. While
$[[i]],[[j]],$ and $[[n]]$ are all metavariables representing natural
numbers, we always use $[[n]]$ to represent term variables that
participate in substitution and $[[i]],[[j]]$ to represent universe levels.
Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable.
Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J c a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[c]]$
is the term whose type is to be casted.
\scw{Example of what you can use identity types for? Or an example of
a program that uses J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}
